---
title: "01 - Download SRP139854 reads"
output: html_document
---

Download the six single-end MBD-BS runs for SRP139854 (*C. virginica* gill, oil vs. no oil),
compress them, record md5 checksums, and build a sample sheet with a `treatment` column.

- Input: `data/runinfo.csv` (SRA run metadata)
- Output reads: `data/raw-reads/` (gitignored)
- Output metadata/checks: `data/sample_sheet.csv`, `output/01-download-data/`

Requires the SRA Toolkit. Create the environment once with:

```
conda env create -f envs/sra-tools.yml
```

```{r setup, include=FALSE}
# Paths are relative to code/, so chunks work the same run one at a time or knitted
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Sample sheet

Sample names starting with `NB` are no-oil controls; `HB` are 25,000 ppm oil.

```{r sample-sheet}
runinfo <- read.csv("../data/runinfo.csv")

sample_sheet <- data.frame(
  run       = runinfo$Run,
  sample    = runinfo$LibraryName,
  treatment = ifelse(grepl("^HB", runinfo$LibraryName), "oil", "control"),
  spots     = runinfo$spots,
  bases     = runinfo$bases
)
sample_sheet <- sample_sheet[order(sample_sheet$treatment, sample_sheet$sample), ]

stopifnot(nrow(sample_sheet) == 6, all(table(sample_sheet$treatment) == 3))

write.csv(sample_sheet, "../data/sample_sheet.csv", row.names = FALSE, quote = FALSE)
sample_sheet
```

# Download

`prefetch` pulls each `.sra` file, then `fasterq-dump` converts it to FASTQ.
Runs that already have a `.fastq.gz` are skipped, so the chunk can be re-run safely.

```{bash download}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/sra-tools/bin:$PATH

threads=8
sra_dir=../data/raw-reads/sra
fq_dir=../data/raw-reads
tmp_dir=${TMPDIR:-/tmp}/fasterq-$$

mkdir -p "${sra_dir}" "${fq_dir}" "${tmp_dir}" ../output/01-download-data

for run in $(tail -n +2 ../data/sample_sheet.csv | cut -d, -f1); do
  if [[ -s "${fq_dir}/${run}.fastq.gz" ]]; then
    echo "${run}: already downloaded, skipping"
    continue
  fi

  prefetch "${run}" \
    --output-directory "${sra_dir}" \
    --max-size 10G

  vdb-validate "${sra_dir}/${run}"

  fasterq-dump "${sra_dir}/${run}" \
    --outdir "${fq_dir}" \
    --temp "${tmp_dir}" \
    --threads "${threads}"

  pigz -p "${threads}" "${fq_dir}/${run}.fastq"
done

rm -rf "${tmp_dir}"
ls -lh "${fq_dir}"
```

# Checksums

```{bash md5}
set -euo pipefail

cd ../data/raw-reads
md5sum SRR*.fastq.gz > ../../output/01-download-data/raw-reads.md5
cat ../../output/01-download-data/raw-reads.md5
```

To verify later: `cd data/raw-reads && md5sum -c ../../output/01-download-data/raw-reads.md5`

# Check read counts against SRA

Every FASTQ should have exactly the number of spots SRA reports.

```{bash read-counts}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/sra-tools/bin:$PATH

echo "run,reads" > ../output/01-download-data/read-counts.csv
for fq in ../data/raw-reads/SRR*.fastq.gz; do
  run=$(basename "${fq}" .fastq.gz)
  n=$(( $(pigz -dc "${fq}" | wc -l) / 4 ))
  echo "${run},${n}" >> ../output/01-download-data/read-counts.csv
done
cat ../output/01-download-data/read-counts.csv
```

```{r compare-counts}
counts <- read.csv("../output/01-download-data/read-counts.csv")
check <- merge(sample_sheet, counts, by = "run")
check$match <- check$spots == check$reads
check[, c("run", "sample", "treatment", "spots", "reads", "match")]

stopifnot(nrow(check) == nrow(sample_sheet), all(check$match))
```
