---
title: "03 - Reference genome and Bismark index"
output: html_document
---

Download the *C. virginica* C_virginica-3.0 assembly (GCF_002022765.2) from NCBI,
verify the files against NCBI's checksums, and build a Bowtie2 Bismark index.

Besides the genome FASTA and GFF, this also grabs two files for step 7:
the GO annotations (`gene_ontology.gaf.gz`) and the RepeatMasker output (`rm.out.gz`, for a TE track).

- Output genome and index: `data/genome/` (gitignored)
- Output checks: `output/03-genome-prep/`

Uses Bismark, Bowtie2 and samtools from the existing `myflow` env.
`envs/bismark.yml` pins the same versions if the env needs to be rebuilt.

```{r setup, include=FALSE}
# Paths are relative to code/, so chunks work the same run one at a time or knitted
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Download

```{bash download}
set -euo pipefail

base=https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0
prefix=GCF_002022765.2_C_virginica-3.0

mkdir -p ../data/genome ../output/03-genome-prep
cd ../data/genome

for suffix in genomic.fna.gz genomic.gff.gz gene_ontology.gaf.gz rm.out.gz; do
  f=${prefix}_${suffix}
  if [[ -s "${f}" ]]; then
    echo "${f}: already downloaded, skipping"
  else
    curl --fail --silent --show-error --retry 3 -O "${base}/${f}"
  fi
done

curl --fail --silent --show-error --retry 3 -o md5checksums.txt "${base}/md5checksums.txt"
ls -lh
```

# Verify checksums

Check each downloaded file against NCBI's `md5checksums.txt`. The chunk fails if any file doesn't match.

```{bash md5}
set -euo pipefail

cd ../data/genome
grep -E "3\.0_(genomic\.fna|genomic\.gff|gene_ontology\.gaf|rm\.out)\.gz$" md5checksums.txt \
  | sed 's#  \./#  #' \
  | md5sum -c - \
  | tee ../../output/03-genome-prep/md5-check.txt
```

# Prepare the FASTA

`bismark_genome_preparation` only picks up `.fa`/`.fasta` files, not NCBI's `.fna`,
so decompress the genome into its own folder with a `.fa` name.

```{bash fasta}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/myflow/bin:$PATH

genome_dir=../data/genome/bismark
fa=${genome_dir}/C_virginica-3.0.fa
mkdir -p "${genome_dir}"

if [[ ! -s "${fa}" ]]; then
  gzip -dc ../data/genome/GCF_002022765.2_C_virginica-3.0_genomic.fna.gz > "${fa}"
fi

samtools faidx "${fa}"
cut -f1,2 "${fa}.fai" > ../output/03-genome-prep/chrom-sizes.tsv
```

```{r genome-summary}
sizes <- read.delim("../output/03-genome-prep/chrom-sizes.tsv", header = FALSE,
                    col.names = c("seqid", "length"))
cat("Sequences:", nrow(sizes), "\n")
cat("Total length (Mb):", round(sum(sizes$length) / 1e6, 1), "\n")
head(sizes[order(-sizes$length), ], 12)
```

# Build the Bismark index

Builds the C->T and G->A converted genomes and indexes each with Bowtie2.
`--parallel 4` runs 4 threads for each of the two indexes (8 cores total).
This takes about an hour, so run it on a compute node.

```{bash bismark-index}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/myflow/bin:$PATH

genome_dir=../data/genome/bismark

if [[ -d "${genome_dir}/Bisulfite_Genome" ]]; then
  echo "Bismark index already exists, skipping"
else
  bismark_genome_preparation \
    --bowtie2 \
    --parallel 4 \
    --verbose \
    "${genome_dir}" \
    2>&1 | tee ../output/03-genome-prep/bismark_genome_preparation.log
fi

ls -R "${genome_dir}/Bisulfite_Genome" | head -30
```
