---
title: "04 - Bismark alignment"
output: html_document
---

Align the trimmed single-end MBD-BS reads to C_virginica-3.0 with Bismark (Bowtie2),
then deduplicate. Both the deduplicated and non-deduplicated BAMs are kept so step 5 can compare them.

Before the full run, a small test on each sample decides whether the library is directional.

- Input: `data/trimmed-reads/SRR*_trimmed.fq.gz` (from `02-qc-trim`), `data/genome/bismark/` (from `03-genome-prep`)
- Output BAMs: `data/bismark/` (gitignored)
- Output reports and summaries: `output/04-bismark-align/`

Uses Bismark, Bowtie2 and samtools from the existing `myflow` env.
Alignment takes a few hours per sample, so run this on a compute node with 20+ cores.

```{r setup, include=FALSE}
# Paths are relative to code/, so chunks work the same run one at a time or knitted
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Is the library directional?

In a directional library, reads only come from the original top and bottom strands (OT, OB).
In a non-directional library (for example Zymo Pico Methyl-Seq), reads also come from their
complements (CTOT, CTOB) in roughly equal numbers.

Align the first 500,000 reads of each sample in `--non_directional` mode, which searches all four strands,
and count how many alignments land on the complementary strands.

```{bash directional-test}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/myflow/bin:$PATH

genome=../data/genome/bismark
out=../output/04-bismark-align/directional-test
mkdir -p "${out}"

for fq in ../data/trimmed-reads/SRR*_trimmed.fq.gz; do
  run=$(basename "${fq}" _trimmed.fq.gz)
  if [[ -s "${out}/${run}_SE_report.txt" ]]; then
    echo "${run}: test already done, skipping"
    continue
  fi
  bismark \
    --genome "${genome}" \
    --non_directional \
    --upto 500000 \
    -p 4 \
    --output_dir "${out}" \
    --temp_dir "${out}" \
    "${fq}"
  # Rename to match the report names used in the full run below
  mv "${out}/${run}_trimmed_bismark_bt2_SE_report.txt" "${out}/${run}_SE_report.txt"
done

# Only the reports are needed from the test
rm -f "${out}"/*.bam
```

```{r directional-summary}
sample_sheet <- read.csv("../data/sample_sheet.csv")

strand_counts <- function(report) {
  lines <- readLines(report)
  get <- function(tag) as.numeric(sub("^[^\t]+\t([0-9]+).*", "\\1", grep(paste0("^", tag, ":\t"), lines, value = TRUE)))
  c(OT = get("CT/CT"), OB = get("CT/GA"), CTOT = get("GA/CT"), CTOB = get("GA/GA"))
}

reports <- list.files("../output/04-bismark-align/directional-test", "_SE_report.txt$", full.names = TRUE)
strands <- as.data.frame(t(sapply(reports, strand_counts)))
strands$run <- sub("_SE_report.txt$", "", basename(reports))
strands$pct_complementary <- round(100 * (strands$CTOT + strands$CTOB) /
                                     rowSums(strands[, c("OT", "OB", "CTOT", "CTOB")]), 2)

strands <- merge(sample_sheet[, c("run", "sample", "treatment")], strands, by = "run")
rownames(strands) <- NULL
strands
```

A directional library shows only a few percent of alignments on CTOT/CTOB (spurious hits);
a non-directional one shows close to 50%. The call below uses 10% as the cutoff.
**Check the table above before trusting it.** Every sample should give the same answer.

```{r directional-call}
is_non_directional <- strands$pct_complementary > 10
stopifnot("Samples disagree on directionality - check the table" =
            all(is_non_directional) || !any(is_non_directional))

library_type <- if (all(is_non_directional)) "non_directional" else "directional"
writeLines(library_type, "../output/04-bismark-align/library-type.txt")
cat("Library type:", library_type, "\n")
```

If the library is non-directional, it is probably Zymo Pico Methyl-Seq.
Consider setting `clip_5`/`clip_3` to 10 in `02-qc-trim` (see the M-bias plots from step 5 to confirm) and re-running from there.

# Align

Reads the library type from the test above. Settings:

- `--parallel 4` runs 4 Bismark instances, each using Bowtie2 with `-p 2`.
  A directional run uses about 16 cores in total; non-directional uses about twice that, so lower `--parallel` if needed.
- `score_min` is Bismark's default (`L,0,-0.2`, about 1 mismatch per 50 bp).
  *C. virginica* is highly polymorphic, so if mapping rates are low, try a looser value such as `L,0,-0.6` and compare.

```{bash align}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/myflow/bin:$PATH

genome=../data/genome/bismark
bam_dir=../data/bismark
report_dir=../output/04-bismark-align/reports
score_min="L,0,-0.2"
mkdir -p "${bam_dir}" "${report_dir}"

library_type=$(cat ../output/04-bismark-align/library-type.txt)
mode_args=()
if [[ "${library_type}" == "non_directional" ]]; then mode_args+=(--non_directional); fi
echo "Library type: ${library_type}"

for fq in ../data/trimmed-reads/SRR*_trimmed.fq.gz; do
  run=$(basename "${fq}" _trimmed.fq.gz)
  if [[ -s "${bam_dir}/${run}.bam" ]]; then
    echo "${run}: already aligned, skipping"
    continue
  fi
  bismark \
    --genome "${genome}" \
    "${mode_args[@]}" \
    --score_min "${score_min}" \
    --parallel 4 \
    -p 2 \
    --output_dir "${bam_dir}" \
    --temp_dir "${bam_dir}" \
    "${fq}"
  # --basename can't be combined with --parallel, so rename Bismark's default output names
  mv "${bam_dir}/${run}_trimmed_bismark_bt2.bam" "${bam_dir}/${run}.bam"
  mv "${bam_dir}/${run}_trimmed_bismark_bt2_SE_report.txt" "${report_dir}/${run}_SE_report.txt"
done

ls -lh "${bam_dir}"
```

# Deduplicate

Removes reads that align to the same position and strand.
MBD enrichment plus single-end reads means some of these "duplicates" may be real independent fragments,
so the original BAMs are kept alongside the deduplicated ones.

```{bash dedup}
set -euo pipefail
export PATH=/mmfs1/gscratch/srlab/sr320/miniforge3/envs/myflow/bin:$PATH

bam_dir=../data/bismark
report_dir=../output/04-bismark-align/reports

for bam in "${bam_dir}"/SRR*.bam; do
  [[ "${bam}" == *.deduplicated.bam ]] && continue
  run=$(basename "${bam}" .bam)
  if [[ -s "${bam_dir}/${run}.deduplicated.bam" ]]; then
    echo "${run}: already deduplicated, skipping"
    continue
  fi
  deduplicate_bismark \
    --single \
    --bam \
    --output_dir "${bam_dir}" \
    "${bam}"
  mv "${bam_dir}/${run}.deduplication_report.txt" "${report_dir}/"
done
```

# Alignment summary

Bisulfite conversion efficiency is estimated from methylation in the CHH context,
which should be close to zero in oysters: conversion ≈ 100 − CHH%.

```{r align-summary}
read_field <- function(lines, pattern) {
  hit <- grep(pattern, lines, value = TRUE)
  if (length(hit) == 0) return(NA)
  as.numeric(sub("^[^\t]*\t\\s*([0-9.]+).*", "\\1", hit[1]))
}

summarize_run <- function(run) {
  align <- readLines(file.path("../output/04-bismark-align/reports", paste0(run, "_SE_report.txt")))
  dedup <- readLines(file.path("../output/04-bismark-align/reports", paste0(run, ".deduplication_report.txt")))
  data.frame(
    run            = run,
    reads          = read_field(align, "^Sequences analysed in total:"),
    unique_aligned = read_field(align, "^Number of alignments with a unique best hit"),
    mapping_pct    = read_field(align, "^Mapping efficiency:"),
    CpG_meth_pct   = read_field(align, "^C methylated in CpG context:"),
    CHG_meth_pct   = read_field(align, "^C methylated in CHG context:"),
    CHH_meth_pct   = read_field(align, "^C methylated in CHH context:"),
    dup_pct        = as.numeric(sub(".*\\(([0-9.]+)%\\).*", "\\1",
                                    grep("^Total number duplicated alignments removed", dedup, value = TRUE)))
  )
}

runs <- sub("_SE_report.txt$", "", list.files("../output/04-bismark-align/reports", "_SE_report.txt$"))
summary_tbl <- do.call(rbind, lapply(runs, summarize_run))
summary_tbl$conversion_pct <- 100 - summary_tbl$CHH_meth_pct
summary_tbl$dedup_aligned <- round(summary_tbl$unique_aligned * (1 - summary_tbl$dup_pct / 100))

summary_tbl <- merge(sample_sheet[, c("run", "sample", "treatment")], summary_tbl, by = "run")
summary_tbl <- summary_tbl[order(summary_tbl$treatment, summary_tbl$sample), ]
rownames(summary_tbl) <- NULL

write.csv(summary_tbl, "../output/04-bismark-align/alignment-summary.csv", row.names = FALSE, quote = FALSE)
summary_tbl
```

Things to check in the table:

- **mapping_pct**: 40–60% is typical for bisulfite reads on oyster. Much lower suggests trying a looser `score_min`.
- **conversion_pct**: should be above 99%. Lower means incomplete bisulfite conversion, which inflates methylation calls.
- **CpG_meth_pct**: should be high for an MBD-enriched library, since the enrichment pulls down methylated DNA.
- **dup_pct** and **dedup_aligned**: HB30 in particular — how many usable reads are left after deduplication.
